Hydra Battles and Cie

### 2021-09-14 doc/chapter-powers.tex
full book with Alectryon (Draft)

We call \emph{chain generator} any function that takes as argument 
any positive integer and returns a chain.

\begin{Coqsrc}
Definition chain_generator := positive -> chain.  
\end{Coqsrc}

A generator $g$  is \emph{correct} it it returns a correct chain
for any exponent:

\begin{Coqsrc}
Definition correct_generator (g : positive -> chain) :=
 forall p, chain_correct (g p) p.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition cpower_pos (g : chain_generator)  p
           `{M:@EMonoid A E_op E_one E_eq} a :=
  chain_apply (g p) (M:=M) a.

Definition cpower (g : chain_generator)  n
           `{M:@EMonoid A E_op E_one E_eq} a :=
  match n with 0%N => E_one 
             | Npos p => cpower_pos  g p a
  end.  
\end{Coqsrc}

@@

Definition optimal_generator (g : positive -> chain) :=
 forall p:positive, optimal p (g p).
\end{Coqsrc}

@@

As for the ``classical'' binary exponentiation algorithm,
we define an auxiliary computation generator for  the
product of an accumulator $a$ with an arbitrary power of some value $x$.
Then, the main function builds a computation for any positive exponent:

\begin{Coqsrc}
Fixpoint axp_scheme  {A} p : A -> A -> @computation A   :=
 match p with
   | xH =>  (fun a x => y <--- a  times x ; Return y)
   | xO q => (fun a x => x2 <--- x times  x ; axp_scheme q a x2)
   | xI q => (fun a x => ax <--- a times x ;
                         x2 <--- x times x ;
                         axp_scheme q ax x2)
end.

Fixpoint  bin_pow_scheme {A} (p:positive)  
: A -> @computation A:=
  match p with 
  |  xH => fun x => Return x
  | xI q  => fun x => x2 <--- x times x; axp_scheme q x x2
  | xO q => fun x => x2 <--- x times x ; bin_pow_scheme q x2
  end.

\end{Coqsrc}

The following function associates  a chain to any positive exponent:

\begin{Coqsrc}
Definition binary_chain (p:positive) : chain :=
  fun A => bin_pow_scheme p.

Compute binary_chain 87.
\end{Coqsrc}

\begin{Coqanswer}
    = fun (A : Type) (x : A) =>
       x0 <--- x times x;
       x1 <--- x times x0;
       x2 <--- x0 times x0;
       x3 <--- x1 times x2;
       x4 <--- x2 times x2;
       x5 <--- x4 times x4;
       x6 <--- x3 times x5;
       x7 <--- x5 times x5;
       x8 <--- x7 times x7; 
       x9 <--- x6 times x8; 
       Return x9
     : chain
\end{Coqanswer}

@@

\begin{Coqsrc}
Section binary_power_proof.

Variables (A: Type)
         (E_op : Mult_op A)
         (E_one : A)
         (E_eq: Equiv A).

Context (M : EMonoid  E_op E_one E_eq).

Existing Instance Eop_proper.
\end{Coqsrc}

\begin{Coqsrc}
Lemma axp_correct : forall p a x,
 computation_eval (axp_scheme p a x) == a *  x ^ (Pos.to_nat p).
(*  Proof by induction on p *)

Lemma binary_correct : 
  forall p x,
        computation_eval  (bin_pow_scheme p (A:=A) x) ==
        x ^ (Pos.to_nat p).
(* Proof by induction on p *)

End  binary_power_proof.
\end{Coqsrc}

\begin{Coqsrc}
Lemma binary_generator_correct : correct_generator binary_chain.
Proof.
  red;unfold chain_correct,  binary_chain, chain_apply;
  split; [auto| intros A op one Eq M x; apply binary_correct].
Qed.
\end{Coqsrc}

@@

It is easy to prove by contradiction  that the binary method is not the most efficient 
for computing powers. 
    First, let us  assume that \texttt{binary\_chain} is optimal:

\begin{Coqsrc}
Section non_optimality_proof.

 Hypothesis binary_opt : optimal binary_chain.
\end{Coqsrc}

Then, let us consider for instance the binary chain generated for the exponent 87.

\begin{Coqsrc}
Compute chain_length (binary_chain 87).
\end{Coqsrc}

\begin{Coqanswer}
   = 10 : nat 
\end{Coqanswer}

Let us recall that \texttt{C87}'s length has been evaluated to $9$ (Sect~\ref{C87-length},
and that this chain is correct (Sect~\vref{C87-param-ok}). Thus, it is very easy
to finish our proof:

\begin{Coqsrc}
Lemma binary_generator_not_optimal : False.
Proof.
  generalize (binary_opt gen _  _ C87_ok); 
  compute; omega.
Qed.

End non_optimality_proof.
\end{Coqsrc}

@@

\begin{Coqsrc}
Example C3 : chain :=
 fun A (x:A) =>
  x2 <--- x times x;
  x3 <--- x2 times x ;
  Return x3.
\end{Coqsrc}

@@

First, we define a type of \emph{continuations},
\emph{i.e.},  functions  that wait for some value $x$, then 
build  a computation for raising {$x$} to some  given exponent.

\begin{Coqsrc}
Definition Fkont (A:Type) := A -> @computation A.
\end{Coqsrc}

An \texttt{f-chain} is just a polymorphic function that combines  a 
continuation and en element into a computation:

\begin{Coqsrc}
Definition Fchain := forall A, Fkont A -> A -> @computation A.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition F3 : Fchain := 
 fun  A k  (x:A) =>
  y <--- x times x ;
  z <--- y times x ;
  k  z.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition F2C (f : Fchain) : chain :=
 fun (A:Type) => f A Return.

Compute the_exponent (F2C F3).
\end{Coqsrc}

\begin{Coqanswer}
= 3%nat  
\end{Coqanswer}

@@

\begin{Coqsrc}
Definition F1 : Fchain := 
 fun A k (x:A) => k x.

Definition F2 : Fchain := 
fun  A k  (x:A) =>
  y <--- x times x ;
  k  y.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition Fapply (f : Fchain) (c: chain) : chain  :=
 fun A  x =>  f  A (fun y => c A  y) x.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition Fcompose (f1 f2: Fchain) : Fchain  :=
 fun   A k x =>  f1  A (fun y => f2 A k y) x.

Lemma F1_neutral_l : forall f, Fcompose F1 f = f.
Proof. reflexivity. Qed.

Lemma F1_neutral_r : forall f, Fcompose f F1 = f.
Proof. reflexivity. Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Example F9 := Fcompose F3 F3.

Compute F9.
\end{Coqsrc}

\begin{Coqanswer}
= fun (A : Type) (x : Fkont A) (x0 : A) =>
       x1 <--- x0 times x0;
       x2 <--- x1 times x0; x3 <--- x2 times x2; 
       x4 <--- x3 times x2; 
       x x4
     : Fchain  
\end{Coqanswer}

\begin{Coqsrc}
Remark F9_correct :chain_correct 9 (F2C F9).
Proof.
  apply param_correctness_pos;  lazy;  parametric_tac.
Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
 Fixpoint  Fexp2_of_nat (n:nat) : Fchain :=
 match n with O => F1
            | S p => Fcompose F2 (Fexp2_of_nat p)
 end.

Definition Fexp2 (p:positive) : Fchain :=
  Fexp2_of_nat (Pos.to_nat p).

Compute Fexp2  4.
\end{Coqsrc}

\begin{Coqanswer}
 = fun (A : Type) (x : Fkont A) (x0 : A) =>
       x1 <--- x0 times x0;
       x2 <--- x1 times x1; x3 <--- x2 times x2; 
       x4 <--- x3 times x3; x x4
     : Fchain
\end{Coqanswer}
%%% ICI

@@

\begin{Coqbad}
Theorem F3_correct : Fchain_correct 3 F3.
Proof.  
  intros    A op E_one E_equiv M k  a ; simpl.
  monoid_simpl M.  
\end{Coqbad}

\begin{Coqanswer}
  A : Type
  op : Mult_op A
  E_one : A
  E_equiv : Equiv A
  M : EMonoid op E_one E_equiv
  k : Fkont A
  a : A
  H : Proper (equiv ==> equiv ==> equiv) op
  ============================
   computation_execute op (k (a * a * a)) ==
   computation_execute op (k (a * (a * (a * E_one))))
\end{Coqanswer}

\begin{Coqbad}
Abort.  
End Bad.
\end{Coqbad}

@@

\begin{Coqsrc}
Class Fkont_proper
      `(M : @EMonoid A op E_one E_equiv) (k: Fkont A )  :=
  Fkont_proper_prf:
    Proper (equiv ==> computation_equiv op E_equiv) k.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition Fchain_correct_nat (n:nat) (f : Fchain) :=
 forall A `(M : @EMonoid A op E_one E_equiv) k
        (Hk :Fkont_proper M k)
        (a : A) ,
 computation_execute op (f A k  a) ==
 computation_execute op (k  (a ^ n)).

Definition Fchain_correct (p:positive) (f : Fchain) :=
 Fchain_correct_nat (Pos.to_nat p) f.
\end{Coqsrc}

@@

Let us show some manual correctness proofs for small f-chains:

\begin{Coqsrc}
Lemma F1_correct : Fchain_correct 1 F1.
Proof.
  intros until M ; intros k Hk a ; unfold F1; simpl.
  apply Hk; monoid_simpl M; reflexivity.
Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Theorem F3_correct : Fchain_correct 3 F3.
Proof. 
  intros until M; intros k Hk a; simpl.
\end{Coqsrc}

\begin{Coqanswer}
 A : Type
  op : Mult_op A
  E_one : A
  E_equiv : Equiv A
  M : EMonoid op E_one E_equiv
  k : Fkont A
  Hk : Fkont_proper M k
  a : A
  ============================
   computation_execute op (k (a * a * a)) ==
   computation_execute op (k (a * (a * (a * E_one))))}
\end{Coqanswer}

\begin{Coqsrc}
apply Hk. 
\end{Coqsrc}

\begin{Coqanswer}
...
============================
a * a * a == a * (a * (a * E_one))}

\begin{Coqsrc}
  monoid_simpl M; reflexivity.
Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Theorem F2_correct : Fchain_correct 2 F2.
Proof. 
  intros until M; intros k Hk a; simpl;
  apply Hk;  monoid_simpl M;  reflexivity.
Qed.
\end{Coqsrc}

@@

We are now looking for a way to generate correct chains for any positive 
number. It seems obvious that we could use \texttt{Fcompose} for building 
a correct f-chain for $n\times p$ by composition of a correct f-chain for 
$n$ and a correct f-chain for $p$.

Let us try to certify this construction:

\begin{Coqbad}
Module Bad2.

Lemma Fcompose_correct_attempt :
  forall f1 f2 n1 n2, Fchain_correct n1 f1 ->
                      Fchain_correct n2 f2 ->
                      Fchain_correct (n1 * n2) (Fcompose f1 f2).

(* Beginning of proof omitted *)
\end{Coqbad}

\begin{Coqanswer}
  Hk : Fkont_proper M k
  a, x, y : A
  Hxy : x == y
  ============================
   computation_execute op (f2 A k x) == 
   computation_execute op (f2 A k y) 
\end{Coqanswer}

@@

\begin{Coqbad}
Abort.
\end{Coqbad}

@@

\begin{Coqbad}
Module Bad3.

Class Fchain_proper_ (fc : Fchain) := Fchain_proper_prf : 
 forall  `(M : @EMonoid A op E_one E_equiv) k  ,
    Fkont_proper M k 
    forall x y, x == y ->
        @computation_equiv _ op E_equiv (fc A k x) (fc A k y).
\end{Coqbad}

@@

\begin{Coqbad}
Instance Fcompose_proper_ (f1 f2 : Fchain)
                             (_ : Fchain_proper_simple f1)
                             (_ : Fchain_proper_simple f2) :
 Fchain_proper_ (Fcompose f1 f2).
Proof. 
 intros until M;intros k Hk x y Hxy; unfold Fcompose;cbn. 
 apply (H _ _ _ _ M); auto.
 intros u v Huv;apply (H0 _ _ _ _ M);auto.
Qed.
\end{Coqbad}

@@

\begin{Coqbad}
End Bad3.
\end{Coqbad}

@@

\begin{Coqsrc}
Definition Fkont_equiv  `(M : @EMonoid A op E_one E_equiv)
 (k k': Fkont A )  := 
 forall x y : A, x == y ->
                 computation_equiv op E_equiv  (k x)  (k' y).

Class Fchain_proper (fc : Fchain) := Fchain_proper_prf : 
 forall  `(M : @EMonoid A op E_one E_equiv) k k' ,
    Fkont_proper M k -> Fkont_proper M k' ->    
    Fkont_equiv M k k' ->
    forall x y, x == y ->
        @computation_equiv _ op E_equiv
                           (fc A k x)
                           (fc A k' y).
\end{Coqsrc}

@@

\begin{Coqsrc}
Instance F1_proper : Fchain_proper F1.
Proof.
  intros until M ; intros k k' Hk Hk' H a b H0; unfold F1; cbn;
  now apply H.  
Qed.
\end{Coqsrc}

\begin{Coqsrc}
Ltac add_op_proper M H := 
 let h := fresh H in
   generalize (@Eop_proper _ _ _ _ M); intro h.

Instance F3_proper : Fchain_proper F3.
Proof.
  intros  A op one equiv M  k k' Hk Hk'  Hkk' x y Hxy;  
  apply Hkk'; add_op_proper M H; repeat rewrite Hxy;
  reflexivity.
Qed.
\end{Coqsrc}

We are now able to prove  \texttt{Fexp2 $n$}'s correctness by induction 
on $n$:

\begin{Coqsrc}
Instance Fexp2_nat_proper (n:nat) : 
                           Fchain_proper (Fexp2_of_nat n).
Proof.
 induction n; cbn.
   - apply F1_proper.
   - apply Fcompose_proper ; [apply F2_proper | apply IHn].
Qed.
\end{Coqsrc}

\begin{Coqsrc}
Lemma  Fexp2_nat_correct (n:nat) : 
         Fchain_correct_nat (2  ^ n) (Fexp2_of_nat n).
Proof.
  induction n; cbn.
 - apply F1_correct.
 -  rewrite Nat.add_0_r;
   replace (2 ^ n + 2 ^ n)%nat with (2 * 2 ^n)%nat by  omega;
   apply Fcompose_correct_nat;auto.
   +  apply F2_correct.
   +  apply  Fexp2_nat_proper.
Qed.
\end{Coqsrc}

\begin{Coqsrc}
Lemma  Fexp2_correct (p:positive) : 
                           Fchain_correct (2  ^ p) (Fexp2 p).
(* Proof omitted *)

Instance  Fexp2_proper (p:positive) : Fchain_proper (Fexp2 p).
(* Proof omitted *)

\end{Coqsrc}

@@

\begin{Coqsrc}
Hint Resolve F1_correct F1_proper
     F3_correct F3_proper Fcompose_correct Fcompose_proper
     Fexp2_correct Fexp2_proper .

Example F144:  {f : Fchain | Fchain_correct 144 f /\
                                Fchain_proper f}.
Proof.
 change 144 with ( (3 * 3) * (2 ^ 4))%positive.
 exists (Fcompose (Fcompose F3 F3) (Fexp2 4)); auto.
Defined.

Compute proj1_sig F144.
\end{Coqsrc}

\begin{Coqanswer}
= fun (A : Type) (x : Fkont A) (x0 : A) =>
       x1 <--- x0 times x0;
       x2 <--- x1 times x0;
       x3 <--- x2 times x2;
       x4 <--- x3 times x2;
       x5 <--- x4 times x4;
       x6 <--- x5 times x5; 
       x7 <--- x6 times x6; 
       x8 <--- x7 times x7; 
       x x8
     : Fchain  
\end{Coqanswer}

@@

\begin{Coqbad}
Definition Fplus (f1 f2 : Fchain) : Fchain :=
  fun A k x => 
   f1 A (fun y => 
            f2 A (fun z => t <--- z times y; k t) x) 
            x.
\end{Coqbad}

For instance, we can define a chain for $23$:

\begin{Coqbad}
Let F23 := Fplus F3 (Fplus (Fexp2 4) (Fexp2 2)).  
\end{Coqbad}

@@

\begin{Coqbad}
Compute F23
\end{Coqbad}

\begin{Coqanswer}
 = fun (A : Type) (k : Fkont A) (x0 : A) =>
       x1 <--- x0 times x0;
       x2 <--- x1 times x0;
       x3 <--- x0 times x0;
       x4 <--- x3 times x3;
       x5 <--- x4 times x4;
       x6 <--- x5 times x5;
       x7 <--- x0 times x0;
       x8 <--- x7 times x7; 
       x9 <--- x8 times x6; 
       x10 <--- x9 times x2; 
       k x10
\end{Coqanswer}

@@

\begin{Coqsrc}
Definition Kkont A:=  A -> A -> @computation A.

Definition Kchain :=  forall A, Kkont A -> A -> @computation A.
\end{Coqsrc}

@@

\begin{Coqsrc}
Example k3_1 : Kchain := fun A (k:Kkont A) (x:A) =>
  x2 <--- x times x ;
  x3 <--- x2 times x ;
  k x3 x.

Example k7_3 : Kchain := fun A (k:Kkont A)  (x:A) =>
  x2 <--- x times x;
  x3 <--- x2 times x ;
  x6 <--- x3 times x3 ;
  x7 <--- x6 times x ;
  k  x7 x3.
\end{Coqsrc}

@@

First, we have to adapt to k-chains our definitions of correctness and properness. 
\begin{Coqsrc}
Definition Kkont_proper `(M : @EMonoid A op E_one E_equiv)
           (k : Kkont A) :=
 Proper (equiv ==> equiv ==> computation_equiv op E_equiv) k .

Definition Kkont_equiv  `(M : @EMonoid A op E_one E_equiv)
           (k k': Kkont A )  := 
 forall x y : A, x == y -> forall z t, z == t -> 
         computation_equiv op E_equiv   (k  x z) (k' y t).
 \end{Coqsrc}

@@

 \begin{Coqsrc}
Definition Kchain_correct_nat (n p : nat) (kc : Kchain) :=
  forall `(M : @EMonoid A op E_one E_equiv)
        (k : Kkont A), 
        Kkont_proper M k ->
        forall  (x : A) ,
              computation_execute op (kc  A k  x) ==
              computation_execute op (k  (x ^ n) (x ^ p)).

\end{Coqsrc}

\begin{Coqsrc}
Definition Kchain_correct (n p : positive) (kc : Kchain) :=
  Kchain_correct_nat  (Pos.to_nat n) (Pos.to_nat p) kc.

Class Kchain_proper (kc : Kchain) :=
Kchain_proper_prf : 
 forall `(M : @EMonoid A op E_one E_equiv) k k' x y ,
   Kkont_proper M k ->
   Kkont_proper M k' -> 
   Kkont_equiv M k k' ->
   E_equiv x y ->
   computation_equiv op E_equiv (kc A k x) (kc A k' y).
\end{Coqsrc}

@@

\begin{Coqsrc}
Instance k7_3_proper : Kchain_proper k7_3.
Proof.
  intros until M; intros; red; unfold k7_3; cbn;
  add_op_proper M H3; apply H1;  rewrite H2;   reflexivity. 
Qed.

Lemma k7_3_correct : Kchain_correct 7 3 k7_3.
Proof.
  intros until M; intros; red; unfold k7_3; simpl.
  apply H; monoid_simpl M;  reflexivity.
Qed. 
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition K2F (knp : Kchain) : Fchain :=
  fun A (k:Fkont A) => kc A (fun  y _ => k y).

Lemma K2F_correct :
  forall knp n p, Kchain_correct kc n p ->
                 Fchain_correct (K2F n) knp.
(* Proof omitted *)

Instance K2F_proper (kc : Kchain)(_ : Kchain_proper kc) :
                                  Fchain_proper (K2F kc).

(* Proof omitted} *)
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition KFK (kbr : Kchain) (fq : Fchain) : Kchain  :=
  fun A k a =>
    kbr A  (fun xb xr =>
              fq A (fun y =>
                      z <--- y times xr; k z xb) xb) a.

Lemma KFK_correct :
  forall (b q r : positive) (kbr : Kchain) (fq : Fchain),
    Kchain_correct b r  kbr ->
    Fchain_correct q fq ->
    Kchain_proper kbr ->
    Fchain_proper fq -> 
    Kchain_correct  (b * q + r) b (KFK kbr fq).
(* Proof omitted *)

Instance KFK_proper :
forall (kbr : Kchain) (fq : Fchain),
  Kchain_proper kbr -> 
  Fchain_proper fq -> 
  Kchain_proper (KFK kbr fq)
(* Proof omitted *)
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition KFF (kbr : Kchain) (fq : Fchain) : Fchain :=
  K2F (KFK kbr fq).

Lemma KFF_correct :
forall (b q r : positive) (kbr : Kchain) (fq : Fchain),
Kchain_correct  b r kbr ->
Fchain_correct q fq ->
Kchain_proper kbr ->
Fchain_proper fq -> Fchain_correct (b * q + r) (KFF kbr fq).
(* Proof omitted *)

Instance KFF_proper :
forall (kbr : Kchain) (fq : Fchain),
Kchain_proper kbr -> Fchain_proper fq -> Fchain_proper (KFF kbr fq).
(* Proof omitted *)
\end{Coqsrc}

@@

Definition FK (f : Fchain) : Kchain :=
  fun (A : Type) (k : Kkont A) (a : A) =>
    f A (fun y => k y a) a.

Lemma FK_correct : forall (p: positive) (Fp : Fchain),
                     Fchain_correct  p Fp ->
                     Fchain_proper Fp ->
                     Kchain_correct p 1 (FK Fp) .
(* Proof omitted *)

Instance  FK_proper  (Fp : Fchain) (_ : Fchain_proper Fp):
  Kchain_proper (FK Fp).
(* Proof omitted *)
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition FFK (fp fq : Fchain) : Kchain :=
  fun A k a =>  fp A (fun xb  => fq A (fun y => k y xb) xb) a.

Lemma FFK_correct  (p q  : positive) (fp fq : Fchain):
    Fchain_correct p fp  ->
    Fchain_correct q fq ->
    Fchain_proper fp ->
    Fchain_proper fq -> Kchain_correct  (p * q ) p (FFK fp fq).
(* Proof omitted *)

Instance FFK_proper  (fp fq : Fchain)
    (_ :   Fchain_proper fp)
    (_ :  Fchain_proper fq) :  Kchain_proper (FFK fp fq) .
(* Proof omitted *)
\end{Coqsrc}

@@

\begin{Coqsrc}
Hint Resolve KFF_correct KFF_proper KFK_correct KFK_proper.

Definition F87 :=
 let k7_3 :=  KFK k3_1 (Fexp2 1) in
 let k10_7 := KFK k7_3 F1 in
 KFF k10_7 (Fexp2 3).

Lemma OK87 : Fchain_correct 87 F87.
Proof.
 unfold F87; change 87 with (10 * (2 ^ 3) + 7)%positive.
 apply KFF_correct;auto.
 change 10 with (7 * 1 + 3); apply KFK_correct;auto.
 change 7 with (3 * 2 ^ 1 + 1)%positive;  apply KFK_correct;auto.
Qed.
\end{Coqsrc}

@@

In the previous section, we  considered two different kinds of objects:
f-chains, associated with a single exponent, and k-chains, associated with two exponents. We would expect that the function \texttt{make\_chain} we want to build and certify is structured as a pair of mutually recursive functions.
 In \coq{} , various ways of building such functions are available:
 \begin{itemize}
 \item Structural [mutual] recursion with \texttt{Fixpoint}
 \item  Using \texttt{Program Fixpoint}
 \item Using   \texttt{Function}.
 \end{itemize}

@@

\begin{Coqsrc}
Inductive signature : Type :=
|  (** Fchain for the exponent n *)
     gen_F (n:positive) 
| (** Kchain for the exponents p+d  and p *) 
     gen_K (p d: positive).
\end{Coqsrc}

@@

\begin{Coqsrc}
(** 
  exponent associated with a signature:
*)
Definition signature_exponent (s:signature) : positive :=
 match s with 
| gen_F n => n 
| gen_K p d  =>  p + d
end.
\end{Coqsrc}

\begin{Coqsrc}
(**
Type of the associated continuation 
*)

Definition kont_type (s: signature)(A:Type) : Type :=
match s with 
| gen_F _  => Fkont A 
| gen_K _ _   => Kkont A
end.

Definition chain_type (s: signature) : Type :=
 match s with 
| gen_F _   => Fchain
|  gen_K _ _  => Kchain
end.

\end{Coqsrc}

\begin{Coqsrc}
Definition correctness_statement (s: signature) : 
chain_type s -> Prop :=
match s  with
  | gen_F p => fun ch => Fchain_correct p ch
  | gen_K p d   => fun ch => Kchain_correct (p + d) p ch
end.

Definition proper_statement (s: signature) : 
chain_type s -> Prop :=
match s  with
  | gen_F p => fun ch => Fchain_proper ch 
  | gen_K p d   => fun ch => Kchain_proper ch 
end.

(**  Full correctness *)

Definition  OK (s: signature) 
  := fun c: chain_type s => 
       correctness_statement s c /\
       proper_statement s c.

\end{Coqsrc}

@@

\vspace{4pt}
\noindent
From ~\href{../theories/html/additions.Dichotomy.html}{additions.Dichotomy}.
\begin{Coqsrc}
Class Strategy (gamma : positive -> positive):=
{
gamma_lt :forall p:positive, 3 < p -> gamma  p < p;
gamma_gt : forall p:positive, 3 < p -> 1 < gamma  p
}.

\end{Coqsrc}
\subsection{The dichotomic strategy}

@@

\begin{Coqsrc}
Function dicho_aux (p:positive) {struct p} : positive :=
 match p with
   | 1%positive   =>  xH
   | 2%positive |   3%positive  => 2 
   | xO (xO q) | xO (xI q) | xI (xO q) | xI (xI q) =>
                                         xO (dicho_aux q)
 end.

Definition dicho  (p:positive) : positive :=
  N2pos (N.div (Npos p) (Npos (dicho_aux p))).

Instance Dicho_strat : Strategy dicho.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition half (p:positive) :=
  match p with xH => xH
          |    xI q | xO q =>  q
  end.

Definition two (p:positive) := 2%positive.

Instance Binary_strat : Strategy half.
Proof.
  split; destruct p; unfold half; try lia.
Qed.

Instance Two_strat : Strategy two.
Proof.
  split;unfold two; lia.
Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition signature_measure (s : signature) : nat :=
match s with
  | gen_F n => 2 * Pos.to_nat n 
  | gen_K p d => 2 * Pos.to_nat (p + d) +1
end.
\end{Coqsrc}

The following function definition generates 9 subgoals,
for proving that the measure on signatures is strictly decreasing along
the recursive calls. They are solved with the help of Standard Library's lemmas 
on arithmetic of \texttt{positive} numbers and Euclidean division.

%\todo{Is Function (from Recdef) obsolete ? Use Equations instead ? }

\begin{Coqsrc}
Function chain_gen  (s:signature) {measure signature_measure}
:  chain_type s :=
  match s  return chain_type s with
    | gen_F i =>
      if pos_eq_dec i 1 then F1 else
        if pos_eq_dec i 3
        then F3
        else 
          match exact_log2 i with
              Some p => Fexp2 p
            | _ =>
              match N.pos_div_eucl i (Npos (dicho i))
              with
                | (q, 0%N) => 
                  Fcompose  (chain_gen (gen_F (dicho i)))
                            (chain_gen (gen_F (N2Pos q)))
                | (q,r)  => KFF (chain_gen
                                   (gen_K (N2Pos r)
                                          (dicho i - N2Pos r)))
                                (chain_gen (gen_F (N2Pos q)))

              end end
\end{Coqsrc}

\begin{Coqsrc}
    | gen_K p d =>
      if pos_eq_dec p 1 then FK (chain_gen (gen_F (1 + d)))
      else
        match N.pos_div_eucl (p + d)  (Npos p) with
          | (q, 0%N) => FFK   (chain_gen (gen_F p))
                              (chain_gen (gen_F (N2Pos q)))
          | (q,r)  => KFK (chain_gen (gen_K (N2Pos r)
                                            (p - N2Pos r)))
                          (chain_gen (gen_F (N2Pos q)))
        end
  end.
(* A lot of arithmetic proofs omitted *)
Defined.

>>>>

We call \emph{chain generator} any function that takes as argument 
any positive integer and returns a chain.
A generator $g$  is \emph{correct} it it returns a correct chain
for any exponent.

\inputsnippets{Addition_Chains/generatorDef}

@@

\inputsnippets{Addition_Chains/cpowerDef}

@@

\inputsnippets{Addition_Chains/optimalGenerator}

@@

As for the ``classical'' binary exponentiation algorithm,
we define an auxiliary computation generator for  the
product of an accumulator $a$ with an arbitrary power of some value $x$.

\inputsnippets{Addition_Chains/binaryChain}

@@

\inputsnippets{Addition_Chains/binaryPowerProofa}
\inputsnippets{Addition_Chains/binaryPowerProofb}
\inputsnippets{Addition_Chains/binaryPowerProofc}
\inputsnippets{Addition_Chains/binaryPowerProofd}
\inputsnippets{Addition_Chains/binaryGeneratorOk}

@@

It is easy to prove by contradiction  that the binary method is not the most efficient 
for computing powers.

\inputsnippets{Addition_Chains/nonOpt}

@@

\inputsnippets{Addition_Chains/C3Example}

@@

First, we define a type of \emph{continuations},
\emph{i.e.},  functions  that wait for some value $x$, then 
build  a computation for raising {$x$} to some  given exponent.
An \texttt{f-chain} is just a polymorphic function that combines  a 
continuation and an element into a computation.

\inputsnippets{Euclidean_Chains/FchainDef}

@@

\inputsnippets{Euclidean_Chains/F3Def}

@@

\inputsnippets{Euclidean_Chains/F2C}

@@

\inputsnippets{Euclidean_Chains/F1F2}

@@

\inputsnippets{Euclidean_Chains/Fapply}

@@

\inputsnippets{Euclidean_Chains/Fcompose}

\inputsnippets{Euclidean_Chains/F1Neutral}

@@

\inputsnippets{Euclidean_Chains/F9Def}
\inputsnippets{Euclidean_Chains/F9Ok}

@@

\inputsnippets{Euclidean_Chains/Fexp2}

@@

\inputsnippets{Euclidean_Chains/BadDefa}
\inputsnippets{Euclidean_Chains/BadDefb}

@@

\inputsnippets{Euclidean_Chains/FkontProper}

@@

\inputsnippets{Euclidean_Chains/GoodFchainCorrect}

@@

Let us show some manual correctness proofs of  small f-chains:

\inputsnippets{Euclidean_Chains/F1Ok}

@@

\inputsnippets{Euclidean_Chains/F3Ok}

@@

\inputsnippets{Euclidean_Chains/F2Ok}

@@

We are now looking for a way to generate correct chains for any positive 
number. It seems obvious that we could use \texttt{Fcompose} for building 
a correct f-chain for $n\times p$ by composition of a correct f-chain for 
$n$ and a correct f-chain for $p$.
Let us try to certify this construction:

\inputsnippets{Euclidean_Chains/Bad2a, Euclidean_Chains/Bad2b}

@@

\inputsnippets{Euclidean_Chains/Bad2c}

@@

\inputsnippets{Euclidean_Chains/Bad3}

@@

\inputsnippets{Euclidean_Chains/Bad3b}

@@

\inputsnippets{Euclidean_Chains/Bad3c}

@@

\inputsnippets{Euclidean_Chains/correctProper}

@@

\inputsnippets{Euclidean_Chains/F1proper}
\inputsnippets{Euclidean_Chains/F2proper}
\inputsnippets{Euclidean_Chains/F3proper}

\subsection{Correctness of chain composition}

The \texttt{Fcompose} operator respects chain correctness and properness.

\inputsnippets{Euclidean_Chains/FcomposeCorrect}
\inputsnippets{Euclidean_Chains/FcomposeProper}

Using chain composition, we get a correct and proper chain for any exponent of the form $2^n$.

\inputsnippets{Euclidean_Chains/Fexp2Correct}
\inputsnippets{Euclidean_Chains/Fexp2Proper}

@@

\inputsnippets{Euclidean_Chains/F144}

@@

\inputsnippets{Euclidean_Chains/Fplus}

@@

\pagebreak
\inputsnippets{Euclidean_Chains/Fplusb}

@@

\inputsnippets{Euclidean_Chains/KchainDef}

@@

\inputsnippets{Euclidean_Chains/K31}
\inputsnippets{Euclidean_Chains/K73}

@@

First, we have to adapt to k-chains our definitions of correctness and properness.

\inputsnippets{Euclidean_Chains/KkontDefs}

@@

\inputsnippets{Euclidean_Chains/KchainCorrectDef}

@@

\inputsnippets{Euclidean_Chains/K73Ok}

@@

\inputsnippets{Euclidean_Chains/K2FDef}
\inputsnippets{Euclidean_Chains/K2FCorrect}
\inputsnippets{Euclidean_Chains/K2FProper}

@@

\inputsnippets{Euclidean_Chains/KFKDef}
\inputsnippets{Euclidean_Chains/KFKCorrect}
\inputsnippets{Euclidean_Chains/KFKProper}

@@

\inputsnippets{Euclidean_Chains/KFFDef}
\inputsnippets{Euclidean_Chains/KFFCorrect}
\inputsnippets{Euclidean_Chains/KFFProper}

@@

\inputsnippets{Euclidean_Chains/FKDef}

Like our other combinators, \texttt{FK} respects chain correctness and properness.

@@

Our last combinator composes a chain for computing $x^p$ with a chain for computing $x^q$ to build a chain for
computing $x^p$ and $x^{pq}$.

\inputsnippets{Euclidean_Chains/FFKDef}
\inputsnippets{Euclidean_Chains/FFKCorrect}
\inputsnippets{Euclidean_Chains/FFKProper}

@@

\inputsnippets{Euclidean_Chains/HintKchains}
\inputsnippets{Euclidean_Chains/F87}
\inputsnippets{Euclidean_Chains/F87Correct}

@@

In the previous section, we  considered two different kinds of objects:
f-chains, associated with a single exponent, and k-chains, associated with two exponents. We would expect that the function \texttt{make\_chain} we want to define and certify is structured as a pair of mutually recursive functions.
 In \coq{} , various ways of building such functions are available:
 \begin{itemize}
 \item Structural [mutual] recursion with \texttt{Fixpoint}
 \item  Using \texttt{Program Fixpoint}
 \item Using   \texttt{Function}.
 \end{itemize}

@@

\inputsnippets{Euclidean_Chains/signature}

@@

\inputsnippets{Euclidean_Chains/dependentlyTypedFuns}

@@

\vspace{4pt}
\noindent
From ~\href{../theories/html/additions.Strategies.html}{additions.Strategies}.

\inputsnippets{Strategies/StrategyDef}

\subsection{The dichotomic strategy}

@@

\inputsnippets{Dichotomy/dichotomy}
\inputsnippets{Dichotomy/DichoStrat}

@@

\inputsnippets{BinaryStrat/BinaryStrats}

@@

\inputsnippets{Euclidean_Chains/GammaContext}

The following function definition generates 9 proof obligations subgoals,
for proving that the measure on signatures is strictly decreasing along
the recursive calls. They are solved with the help of Standard Library's lemmas 
on arithmetic of \texttt{positive} numbers and Euclidean division.

\pagebreak
\inputsnippets{Euclidean_Chains/chainGen}
\inputsnippets{Euclidean_Chains/makeChain}
\inputsnippets{Euclidean_Chains/C87Dicho}
